### 2025-03-14 paper/quarto_ecml/paper.tex
Back to (exactly) 16 pages!
TODO: Complete read-through

In the natural language processing (NLP) domain, CEs have also been used
to improve models through data augmentation. \cite{wu2021polyjuice}
propose \emph{Polyjuice}, a general-purpose counterfactual generator for
language models. The authors empirically demonstrate that the
augmentation of training data with their method improves robustness in a
number of NLP tasks. \cite{balashankar2023improving} similarly use
\emph{Polyjuice} to augment NLP datasets through diverse counterfactuals
and show that classifier robustness improves by up to 20\%. Finally,
\cite{luu2023counterfactual} introduce Counterfactual Adversarial
Training (CAT), which also aims at improving generalization and
robustness of language models through a three-step procedure: the
authors identify training samples that are subject to high predictive
uncertainty, generate CEs for those samples, and then fine-tune the
language model on a dataset augmented with the CEs.

@@

The characterization of faithfulness and plausibility in Def.
\ref{def-explainability} is the same as in \cite{altmeyer2024faithful},
with adapted notation. Intuitively, plausible counterfactuals are
consistent with the data and faithful counterfactuals are consistent
with what the model has learned about input data. Actionability
constraints in Def. \ref{def-explainability} vary and depend on the
context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
focus on domain and mutability constraints for individual features
\(x_d\) for \(d=1,...,D\). We limit ourselves to classification tasks
for reasons discussed in Section~\ref{sec-discussion}.

@@

As observed by \cite{grathwohl2020your}, any classifier can be
re-interpreted as a joint energy-based model (JEM) that learns to
discriminate output classes conditional on the observed (training)
samples from \(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
perform well at both tasks by directly maximizing the joint
log-likelihood factorized as
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
The first term can be optimized using conventional cross-entropy as in
Equation~\ref{eq-obj}. Then, to optimize \(\log p_\theta(\mathbf{x})\)
\cite{grathwohl2020your} minimizes the contrastive divergence between
these observed samples from \(p(\mathbf{x})\) and generated samples from
\(p_\theta(\mathbf{x})\).

@@

From a Bayesian perspective, setting
\(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) can be understood as
assuming a point mass prior for \(p(\mathbf{x}^+)\) with respect to
feature \(d\). Intuitively, we think of this simply in terms ignoring
implausibility costs with respect to immutable features, which
effectively forces the model to instead seek plausibility with respect
to the remaining features. This in turn results in lower overall
sensitivity to immutable features, which we demonstrate empirically for
different classifiers in Section~\ref{sec-experiments}. Under certain
conditions, this result holds theoretically:\footnote{For the proof, see
  the supplementary appendix.}

@@

It is worth highlighting that \mbox{Proposition \ref{prp-mtblty}}
assumes independence of features. This raises a valid concern about the
effect of protecting immutable features in the presence of proxies that
remain unprotected. We address this in Section~\ref{sec-discussion}.

@@

These broad and substantial improvements in plausibility generally do
not come at the cost of decreased predictive performance: test accuracy
for CT is virtually identical to the baseline for \emph{Adult},
\emph{Circ}, \emph{LS}, \emph{Moon} and \emph{OL}, and even slightly
improved for \emph{Cred}. Exceptions to this general pattern are
\emph{MNIST}, \emph{CH} and \emph{GMSC}, for which we observe reduction
in test accuracy of 2, 5 and 15 percentage points, respectively. We note
in this context, that we have not optimized our models for predictive
performance at all and worked with very small networks. In summary, we
find that CT can substantially improve the quality of explanations
learned by models without sacrificing predictive accuracy.

@@

\caption{\label{tbl-main}Key explainability and predictive performance
metrics for all datasets. \textbf{Plausibility}: the top five rows show
the percentage reduction in implausibility according to
Equation~\ref{eq-impl-dist} for varying degrees of the energy penalty
used for \emph{ECCo} at test time. The following row shows the reduction
in implausibility as measured by Equation~\ref{eq-impl-div} and
aggregated across all test specifications. \textbf{Accuracy}: The
following two rows show the test accuracies for the models trained with
CT and the baseline. \textbf{Actionability}: The final row present the
average reduction in costs when imposing mutability constraints.}

@@

In Section~\ref{sec-method}, we show that our proposed way for encoding
mutability constraints leads to lower classifier sensitivity wrt.
immutable features for linear models, tilting the decision boundary in
favour of mutable features instead. For binding constraints at test
time, this has the effect of shorter counterfactual paths and hence
smaller average costs (\(\ell_1\)-norm) to individuals. To extend this
to the non-linear case, we test the effect of imposing mutability
constraints empirically for our synthetic data using the same evaluation
scheme as above. The final row in Table~\ref{tbl-main} reports the
average reduction in costs for CT compared to the baseline, when
imposing that either the first or the second feature is immutable. In
all cases, costs are reduced substantially, indicating that classifiers
trained with CT are indeed more sensitive to mutable features.

@@

We test the impact of three types of hyperparameters; our complete
results are in the supplementary appendix.\\
\indent \emph{Hyperparameters of the CE generators.} First, we observe
that CT is highly sensitive to hyperparameter settings but (a) there are
manageable patterns and (b) we can typically identify settings that
improve either plausibility or cost, and commonly both of them at the
same time. Second, we note that the choice of a CE generator has a major
impact on the results. For example, \emph{REVISE} tends to perform the
worst, most likely because it uses a surrogate VAE to generate
counterfactuals which impedes faithfulness \cite{altmeyer2024faithful}.
Third, increasing \(T\), the maximum number of steps, generally yields
better outcomes because more CEs can mature in each training epoch.
Fourth, the impact of \(\tau\), the required decision threshold is more
difficult to predict. On ``harder'' datasets it may be difficult to
satisfy high \(\tau\) for any given sample (i.e., also factuals) and so
increasing this threshold does not seem to correlate with better
outcomes. In fact, we have generally found that a choice of \(\tau=0.5\)
leads to optimal results because it is associated with high proportions
of mature counterfactuals.\\
\indent \emph{Hyperparameters for penalties.} We find that the strength
of the energy regularization, \(\lambda_{\text{reg}}\) is highly
impactful; energy must be sufficiently regularized to avoid poor
performance in terms of decreased plausibility and increased costs. The
sensitivity with respect to \(\lambda_{\text{div}}\) and
\(\lambda_{\text{adv}}\) is much less evident. While high values of
\(\lambda_{\text{reg}}\) may increase the variability in outcomes
(especially) when combined with high values of \(\lambda_{\text{div}}\)
or \(\lambda_{\text{adv}}\), this effect is not very pronounced.\\
\indent \emph{Other hyperparameters.} We observe that the effectiveness
and stability of CT is positively associated with the number of
counterfactuals generated during each training epoch. We also confirm
that a higher number of training epochs is beneficial. Interestingly, we
observed desired improvements in explainability when CT was combined
with conventional training and applied only for the final 50\% of epochs
of the complete training process. Put differently, CT may be a way to
improve the explainability of models in a fine-tuning manner.

@@

We also highlight several important directions for future research.
Firstly, it is an interesting challenge to extend CT beyond
classification settings. Our formulation relies on the distinction
between non-target class(es) \(y^{-}\) and target class(es) \(y^{+}\) to
generate counterfactuals through Equation~\ref{eq-obj}. While \(y^{-}\)
and \(y^{+}\) can be arbitrarily defined, CT requires the output space
\(\mathcal{Y}\) to be discrete. Thus, it does not apply to ML tasks
where the change in outcome cannot be readily quantified. Focus on
classification models is a common restriction in research on CEs and AR.
Other settings have attracted some interest (e.g., regression in
\cite{spooner2021counterfactual,zhao2023counterfactual}), but there is
little consensus how to robustly extend the notion of CEs.\\
\indent Secondly, our approach is susceptible to training instabilities.
This problem has been recognized for JEMs \cite{grathwohl2020your} and
even though we depart from the SGLD-based sampling, we still encounter
considerable variability in the outcomes. CT is exposed to two potential
sources of instabilities: (1) the energy-based contrastive divergence
term in Equation~\ref{eq-div}, and (2) the underlying counterfactual
explainers. We find several promising ways to mitigate this problem:
regularizing energy (\(\lambda_{\text{reg}}\)), generating sufficiently
many counterfactuals during each epoch, and including only mature
counterfactuals for contrastive divergence.\\
\indent Finally, we believe that it is possible to substantially improve
hyperparameter selection procedures. Our method benefits from the tuning
of certain key hyperparameters (see Section~\ref{sec-hyperparameters}).
In this work, we have relied exclusively on grid search for this task.
Future work on CT could benefit from investigating more sophisticated
approaches. Notably, CT is iterative which makes methods such as
Bayesian (e.g., \cite{snoek2012practical}) or gradient-based (e.g.,
\cite{franceschi2017forward}) optimization applicable.

>>>>

In the natural language processing (NLP) domain, CEs have also been used
to improve models through data augmentation. \cite{wu2021polyjuice2}
propose \emph{Polyjuice}, a general-purpose counterfactual generator for
language models. The authors empirically demonstrate that the
augmentation of training data with their method improves robustness in a
number of NLP tasks. \cite{balashankar2023improving} similarly use
\emph{Polyjuice} to augment NLP datasets through diverse counterfactuals
and show that classifier robustness improves by up to 20\%. Finally,
\cite{luu2023counterfactual} introduce Counterfactual Adversarial
Training (CAT), which also aims at improving generalization and
robustness of language models through a three-step procedure: the
authors identify training samples that are subject to high predictive
uncertainty, generate CEs for those samples, and then fine-tune the
language model on a dataset augmented with the CEs.

@@

\noindent The characterization of faithfulness and plausibility in Def.
\ref{def-explainability} is the same as in \cite{altmeyer2024faithful},
with adapted notation. Intuitively, plausible counterfactuals are
consistent with the data and faithful counterfactuals are consistent
with what the model has learned about input data. Actionability
constraints in Def. \ref{def-explainability} vary and depend on the
context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
focus on domain and mutability constraints for individual features
\(x_d\) for \(d=1,...,D\). We limit ourselves to classification tasks
for reasons discussed in Section~\ref{sec-discussion}.

@@

As observed by \cite{grathwohl2020your}, any classifier can be
re-interpreted as a joint energy-based model (JEM) that learns to
discriminate output classes conditional on the observed (training)
samples from \(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
perform well at both tasks by directly maximizing the joint
log-likelihood factorized as
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
The first term can be optimized using conventional cross-entropy as in
Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\)
\cite{grathwohl2020your} minimizes the contrastive divergence between
the observed samples from \(p(\mathbf{x})\) and generated samples from
\(p_\theta(\mathbf{x})\).

@@

\noindent From a Bayesian perspective, setting
\(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) can be understood as
assuming a point mass prior for \(p(\mathbf{x}^+)\) with respect to
feature \(d\). Intuitively, we think of this simply in terms ignoring
implausibility costs with respect to immutable features, which
effectively forces the model to instead seek plausibility with respect
to the remaining features. This in turn results in lower overall
sensitivity to immutable features, which we demonstrate empirically for
different classifiers in Section~\ref{sec-experiments}. Under certain
conditions, this result holds theoretically:\footnote{For the proof, see
  the supplementary appendix.}

@@

\noindent It is worth highlighting that
\mbox{Proposition \ref{prp-mtblty}} assumes independence of features.
This raises a valid concern about the effect of protecting immutable
features in the presence of proxies that remain unprotected. We address
this in Section~\ref{sec-discussion}.

@@

We find that CT can substantially improve the quality of explanations
learned by models without sacrificing predictive performance. Test
accuracy for CT is virtually identical to the baseline for \emph{Adult},
\emph{Circ}, \emph{LS}, \emph{Moon} and \emph{OL}, and even slightly
improved for \emph{Cred}. Exceptions to this general pattern are
\emph{MNIST}, \emph{CH} and \emph{GMSC}, for which we observe reduction
in test accuracy of 2, 5 and 15 percentage points, respectively. In this
context, we also note that we have not optimized the models (very small
networks) for predictive performance at all.

@@

\caption{\label{tbl-main}Key performance metrics across all datasets.
\textbf{Plausibility}: Rows 1-5 show the percentage reduction in
implausibility via Equation~\ref{eq-impl-dist} for varying degrees of
the energy penalty used for \emph{ECCo} at test time; row 6 shows the
reduction in implausibility via Equation~\ref{eq-impl-div} and
aggregated across all test specifications. \textbf{Accuracy} (rows 7-8):
test accuracies for models trained with CT and the baseline.
\textbf{Actionability} (row 9): average reduction in costs when imposing
mutability constraints.}

@@

In Section~\ref{sec-method}, we show that our proposed way for encoding
mutability constraints leads to lower classifier sensitivity wrt.
immutable features for linear models, tilting the decision boundary in
favour of mutable features instead. For binding constraints at test
time, this leads to shorter counterfactual paths and hence smaller
average costs (\(\ell_1\)-norm) to individuals. To extend this to the
non-linear case, we test the effect of imposing mutability constraints
empirically for our synthetic data using the same evaluation scheme as
above. The final row in Table~\ref{tbl-main} reports the average
reduction in costs for CT compared to the ``vanilla'' baseline, when
imposing that either the first or the second feature is immutable. In
all cases, costs are reduced substantially, indicating that classifiers
trained with CT are indeed more sensitive to mutable features.

@@

We test the impact of three types of hyperparameters; our complete
results are in the supplementary appendix.\\
\indent We note that CT is highly sensitive to the choice of a CE
generator and its hyperparameters but (a) there are manageable patterns
and (b) we can typically identify settings that improve either
plausibility or cost, and commonly both of them at the same time. For
example, \emph{REVISE} tends to perform the worst, most likely because
it uses a surrogate VAE to generate counterfactuals which impedes
faithfulness \cite{altmeyer2024faithful}. Increasing \(T\), the maximum
number of steps, generally yields better outcomes because more CEs can
mature in each training epoch. The impact of \(\tau\), the required
decision threshold is more difficult to predict. On ``harder'' datasets
it may be difficult to satisfy high \(\tau\) for any given sample (i.e.,
also factuals) and so increasing this threshold does not seem to
correlate with better outcomes. In fact, the choice of \(\tau=0.5\)
generally leads to optimal results because it is associated with high
proportions of mature counterfactuals.\\
\indent The strength of the energy regularization,
\(\lambda_{\text{reg}}\) is highly impactful and leads to poor
performance in terms of decreased plausibility and increased costs if
insufficiently high. The sensitivity with respect to
\(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\) is much less
evident. While high values of \(\lambda_{\text{reg}}\) may increase the
variability in outcomes when combined with high values of
\(\lambda_{\text{div}}\) or \(\lambda_{\text{adv}}\), this effect is not
very pronounced.\\
\indent The effectiveness and stability of CT is positively associated
with the number of counterfactuals generated during each training epoch.
We also confirm that a higher number of training epochs is beneficial.
Interestingly, we observed desired improvements when CT was combined
with conventional training and applied only for the final 50\% of epochs
of the complete training process. Put differently, CT can improve the
explainability of models in a fine-tuning manner.

@@

We also highlight several important directions for future research.
Firstly, it is an interesting challenge to extend CT beyond
classification settings. Our formulation relies on the distinction
between non-target class(es) \(y^{-}\) and target class(es) \(y^{+}\) to
generate counterfactuals through Equation~\ref{eq-obj}. While \(y^{-}\)
and \(y^{+}\) can be arbitrarily defined, CT requires the output space
\(\mathcal{Y}\) to be discrete. Thus, it does not apply to ML tasks
where the change in outcome cannot be readily quantified. Focus on
classification models is a common restriction in research on CEs and AR.
Other settings have attracted some interest (e.g., regression in
\cite{spooner2021counterfactual}), but there is little consensus how to
robustly extend the notion of CEs.\\
\indent Secondly, our approach is susceptible to training instabilities.
This problem has been recognized for JEMs \cite{grathwohl2020your} and
even though we depart from the SGLD-based sampling, we still encounter
considerable variability in the outcomes. CT is exposed to two potential
sources of instabilities: (1) the energy-based contrastive divergence
term in Equation~\ref{eq-div}, and (2) the underlying counterfactual
explainers. We find several promising ways to mitigate this problem:
regularizing energy (\(\lambda_{\text{reg}}\)), generating sufficiently
many counterfactuals during each epoch, and including only mature
counterfactuals for contrastive divergence.\\
\indent Finally, we believe that it is possible to substantially improve
hyperparameter selection procedures. Our method benefits from the tuning
of certain key hyperparameters (see Section~\ref{sec-hyperparameters}).
In this work, we have relied exclusively on grid search for this task.
Future work on CT could benefit from investigating more sophisticated
approaches. Notably, CT is iterative which makes methods such as
Bayesian or gradient-based optimization applicable (see
\cite{bischl2023hyperparameter}).

### 2025-03-14 paper/sections/discussion.qmd
Back to (exactly) 16 pages!
TODO: Complete read-through

As our results indicate, counterfactual training produces models that are more explainable. Nonetheless, it brings about three important limitations.  
`\indent`{=latex} *CT increases the training time of models.* CT can be more time-consuming than conventional training regimes. While higher numbers of CEs per iteration positively impact the quality of solutions, they also increase the amount of computations. Relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs.^[See supplementary appendix for computational details.] Three factors attenuate this effect. First, CT amortizes the cost of CEs for the training samples. Second, we find that it can retain its value when used as a "fine-tuning" technique for conventionally-trained models. Third, it yields itself to parallel execution, which we have leveraged for our own experiments.  
`\indent`{=latex} *Immutable features may have proxies.* We propose an approach to protect immutable features and thus increase the actionability of the generated CEs. However, it requires that model owners define the mutability constraints for (all) features considered by the model. Even if all immutable features are protected, there may exist proxies that are mutable (and hence should not be protected) but preserve enough information about the principals to hinder the protections. Delineating actionability is a major undecided challenge in the AR literature [see, e.g., @venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal.  
`\indent`{=latex} *Interventions on features may impact fairness.* We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of decision subjects. As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. For example, consider the Adult dataset used in our experiments, where *workclass* or *education* may be more difficult to change for underpriviledged groups. When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [e.g., @sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness]. Still, these phenomena are not specific to CT.

We also highlight several important directions for future research. Firstly, it is an interesting challenge to extend CT beyond classification settings. Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj. While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete. Thus, it does not apply to ML tasks where the change in outcome cannot be readily quantified. Focus on classification models is a common restriction in research on CEs and AR. Other settings have attracted some interest (e.g., regression in [@spooner2021counterfactual; @zhao2023counterfactual]), but there is little consensus how to robustly extend the notion of CEs.  
`\indent`{=latex} Secondly, our approach is susceptible to training instabilities. This problem has been recognized for JEMs [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter considerable variability in the outcomes. CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers. We find several promising ways to mitigate this problem: regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each epoch, and including only mature counterfactuals for contrastive divergence.  
`\indent`{=latex} Finally, we believe that it is possible to substantially improve hyperparameter selection procedures. Our method benefits from the tuning of certain key hyperparameters (see @sec-hyperparameters). In this work, we have relied exclusively on grid search for this task. Future work on CT could benefit from investigating more sophisticated approaches. Notably, CT is iterative which makes methods such as Bayesian (e.g., [@snoek2012practical]) or gradient-based (e.g., [@franceschi2017forward]) optimization applicable.

>>>>

As our results indicate, counterfactual training produces models that are more explainable. Nonetheless, it brings about three important limitations.  
`\indent`{=latex} *CT increases the training time of models.* CT can be more time-consuming than conventional training regimes. While higher numbers of CEs per iteration positively impact the quality of solutions, they also increase the amount of computations. Relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs.^[See supplementary appendix for computational details.] Three factors attenuate this effect. First, CT amortizes the cost of CEs for the training samples. Second, we find that it can retain its value when used as a "fine-tuning" technique for conventionally-trained models. Third, it yields itself to parallel execution, which we have leveraged for our own experiments.  
`\indent`{=latex} *Immutable features may have proxies.* We propose an approach to protect immutable features and thus increase the actionability of the generated CEs. However, it requires that model owners define the mutability constraints for (all) features considered by the model. Even if all immutable features are protected, there may exist proxies that are mutable (and hence should not be protected) but preserve enough information about the principals to hinder the protections. Delineating actionability is a major undecided challenge in the AR literature [see, e.g., @venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal.  
`\indent`{=latex} *Interventions on features may impact fairness.* We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of decision subjects. As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. For example, consider the Adult dataset used in our experiments, where *workclass* or *education* may be more difficult to change for underpriviledged groups. When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [@sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness]. Still, these phenomena are not specific to CT.

We also highlight several important directions for future research. Firstly, it is an interesting challenge to extend CT beyond classification settings. Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj. While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete. Thus, it does not apply to ML tasks where the change in outcome cannot be readily quantified. Focus on classification models is a common restriction in research on CEs and AR. Other settings have attracted some interest (e.g., regression in [@spooner2021counterfactual]), but there is little consensus how to robustly extend the notion of CEs.  
`\indent`{=latex} Secondly, our approach is susceptible to training instabilities. This problem has been recognized for JEMs [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter considerable variability in the outcomes. CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers. We find several promising ways to mitigate this problem: regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each epoch, and including only mature counterfactuals for contrastive divergence.  
`\indent`{=latex} Finally, we believe that it is possible to substantially improve hyperparameter selection procedures. Our method benefits from the tuning of certain key hyperparameters (see @sec-hyperparameters). In this work, we have relied exclusively on grid search for this task. Future work on CT could benefit from investigating more sophisticated approaches. Notably, CT is iterative which makes methods such as Bayesian or gradient-based optimization applicable (see [@bischl2023hyperparameter]).

### 2025-03-14 paper/sections/experiments.qmd
Back to (exactly) 16 pages!
TODO: Complete read-through

These broad and substantial improvements in plausibility generally do not come at the cost of decreased predictive performance: test accuracy for CT is virtually identical to the baseline for *Adult*, *Circ*, *LS*, *Moon* and *OL*, and even slightly improved for *Cred*. Exceptions to this general pattern are *MNIST*, *CH* and *GMSC*, for which we observe reduction in test accuracy of 2, 5 and 15 percentage points, respectively. We note in this context, that we have not optimized our models for predictive performance at all and worked with very small networks. In summary, we find that CT can substantially improve the quality of explanations learned by models without sacrificing predictive accuracy.

@@

Key explainability and predictive performance metrics for all datasets. **Plausibility**: the top five rows show the percentage reduction in implausibility according to @eq-impl-dist for varying degrees of the energy penalty used for *ECCo* at test time. The following row shows the reduction in implausibility as measured by @eq-impl-div and aggregated across all test specifications. **Accuracy**: The following two rows show the test accuracies for the models trained with CT and the baseline. **Actionability**: The final row present the average reduction in costs when imposing mutability constraints.

:::

@@

In @sec-method, we show that our proposed way for encoding mutability constraints leads to lower classifier sensitivity wrt. immutable features for linear models, tilting the decision boundary in favour of mutable features instead. For binding constraints at test time, this has the effect of shorter counterfactual paths and hence smaller average costs ($\ell_1$-norm) to individuals. To extend this to the non-linear case, we test the effect of imposing mutability constraints empirically for our synthetic data using the same evaluation scheme as above. The final row in @tbl-main reports the average reduction in costs for CT compared to the baseline, when imposing that either the first or the second feature is immutable. In all cases, costs are reduced substantially, indicating that classifiers trained with CT are indeed more sensitive to mutable features.

### Impact of hyperparameter settings. {#sec-hyperparameters}
We test the impact of three types of hyperparameters; our complete results are in the supplementary appendix.  
`\indent`{=latex} *Hyperparameters of the CE generators.* First, we observe that CT is highly sensitive to hyperparameter settings but (a) there are manageable patterns and (b) we can typically identify settings that improve either plausibility or cost, and commonly both of them at the same time. Second, we note that the choice of a CE generator has a major impact on the results. For example, *REVISE* tends to perform the worst, most likely because it uses a surrogate VAE to generate counterfactuals which impedes faithfulness [@altmeyer2024faithful]. Third, increasing $T$, the maximum number of steps, generally yields better outcomes because more CEs can mature in each training epoch. Fourth, the impact of $\tau$, the required decision threshold is more difficult to predict. On "harder" datasets it may be difficult to satisfy high $\tau$ for any given sample (i.e., also factuals) and so increasing this threshold does not seem to correlate with better outcomes. In fact, we have generally found that a choice of $\tau=0.5$ leads to optimal results because it is associated with high proportions of mature counterfactuals.  
`\indent`{=latex} *Hyperparameters for penalties.* We find that the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful; energy must be sufficiently regularized to avoid poor performance in terms of decreased plausibility and increased costs. The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident. While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes (especially) when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not very pronounced.   
`\indent`{=latex} *Other hyperparameters.* We observe that the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch. We also confirm that a higher number of training epochs is beneficial. Interestingly, we observed desired improvements in explainability when CT was combined with conventional training and applied only for the final 50% of epochs of the complete training process. Put differently, CT may be a way to improve the explainability of models in a fine-tuning manner.

>>>>

We find that CT can substantially improve the quality of explanations learned by models without sacrificing predictive performance. Test accuracy for CT is virtually identical to the baseline for *Adult*, *Circ*, *LS*, *Moon* and *OL*, and even slightly improved for *Cred*. Exceptions to this general pattern are *MNIST*, *CH* and *GMSC*, for which we observe reduction in test accuracy of 2, 5 and 15 percentage points, respectively. In this context, we also note that we have not optimized the models (very small networks) for predictive performance at all.

@@

Key performance metrics across all datasets. **Plausibility**: Rows 1-5 show the percentage reduction in implausibility via @eq-impl-dist for varying degrees of the energy penalty used for *ECCo* at test time; row 6 shows the reduction in implausibility via @eq-impl-div and aggregated across all test specifications. **Accuracy** (rows 7-8): test accuracies for models trained with CT and the baseline. **Actionability** (row 9): average reduction in costs when imposing mutability constraints.
:::

@@

In @sec-method, we show that our proposed way for encoding mutability constraints leads to lower classifier sensitivity wrt. immutable features for linear models, tilting the decision boundary in favour of mutable features instead. For binding constraints at test time, this leads to shorter counterfactual paths and hence smaller average costs ($\ell_1$-norm) to individuals. To extend this to the non-linear case, we test the effect of imposing mutability constraints empirically for our synthetic data using the same evaluation scheme as above. The final row in @tbl-main reports the average reduction in costs for CT compared to the "vanilla" baseline, when imposing that either the first or the second feature is immutable. In all cases, costs are reduced substantially, indicating that classifiers trained with CT are indeed more sensitive to mutable features.

### Impact of hyperparameter settings. {#sec-hyperparameters}
We test the impact of three types of hyperparameters; our complete results are in the supplementary appendix.  
`\indent`{=latex} We note that CT is highly sensitive to the choice of a CE generator and its hyperparameters but (a) there are manageable patterns and (b) we can typically identify settings that improve either plausibility or cost, and commonly both of them at the same time. For example, *REVISE* tends to perform the worst, most likely because it uses a surrogate VAE to generate counterfactuals which impedes faithfulness [@altmeyer2024faithful]. Increasing $T$, the maximum number of steps, generally yields better outcomes because more CEs can mature in each training epoch. The impact of $\tau$, the required decision threshold is more difficult to predict. On "harder" datasets it may be difficult to satisfy high $\tau$ for any given sample (i.e., also factuals) and so increasing this threshold does not seem to correlate with better outcomes. In fact, the choice of $\tau=0.5$ generally leads to optimal results because it is associated with high proportions of mature counterfactuals.  
`\indent`{=latex} The strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful and leads to poor performance in terms of decreased plausibility and increased costs if insufficiently high. The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident. While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not very pronounced.   
`\indent`{=latex} The effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch. We also confirm that a higher number of training epochs is beneficial. Interestingly, we observed desired improvements when CT was combined with conventional training and applied only for the final 50% of epochs of the complete training process. Put differently, CT can improve the explainability of models in a fine-tuning manner.

### 2025-03-14 paper/sections/method.qmd
Back to (exactly) 16 pages!
TODO: Complete read-through

The characterization of faithfulness and plausibility in Def. \ref{def-explainability} is the same as in @altmeyer2024faithful, with adapted notation. Intuitively, plausible counterfactuals are consistent with the data and faithful counterfactuals are consistent with what the model has learned about input data. Actionability constraints in Def. \ref{def-explainability} vary and depend on the context in which $\mathbf{M}_\theta$ is deployed. In this work, we focus on domain and mutability constraints for individual features $x_d$ for $d=1,...,D$. We limit ourselves to classification tasks for reasons discussed in @sec-discussion.

@@

As observed by @grathwohl2020your, any classifier can be re-interpreted as a joint energy-based model (JEM) that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$. The authors show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood factorized as $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$. The first term can be optimized using conventional cross-entropy as in @eq-obj. Then, to optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimizes the contrastive divergence between these observed samples from $p(\mathbf{x})$ and generated samples from $p_\theta(\mathbf{x})$.

@@

From a Bayesian perspective, setting $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x}^+)$ with respect to feature $d$. Intuitively, we think of this simply in terms ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features. This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically for different classifiers in @sec-experiments. Under certain conditions, this result holds theoretically:^[For the proof, see the supplementary appendix.]

@@

It is worth highlighting that \mbox{Proposition \ref{prp-mtblty}} assumes independence of features. This raises a valid concern about the effect of protecting immutable features in the presence of proxies that remain unprotected. We address this in @sec-discussion.

>>>>

`\noindent`{=latex} The characterization of faithfulness and plausibility in Def. \ref{def-explainability} is the same as in @altmeyer2024faithful, with adapted notation. Intuitively, plausible counterfactuals are consistent with the data and faithful counterfactuals are consistent with what the model has learned about input data. Actionability constraints in Def. \ref{def-explainability} vary and depend on the context in which $\mathbf{M}_\theta$ is deployed. In this work, we focus on domain and mutability constraints for individual features $x_d$ for $d=1,...,D$. We limit ourselves to classification tasks for reasons discussed in @sec-discussion.

@@

As observed by @grathwohl2020your, any classifier can be re-interpreted as a joint energy-based model (JEM) that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$. The authors show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood factorized as $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$. The first term can be optimized using conventional cross-entropy as in @eq-obj. To optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimizes the contrastive divergence between the observed samples from $p(\mathbf{x})$ and generated samples from $p_\theta(\mathbf{x})$.

@@

`\noindent`{=latex} From a Bayesian perspective, setting $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x}^+)$ with respect to feature $d$. Intuitively, we think of this simply in terms ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features. This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically for different classifiers in @sec-experiments. Under certain conditions, this result holds theoretically:^[For the proof, see the supplementary appendix.]

@@

`\noindent`{=latex} It is worth highlighting that \mbox{Proposition \ref{prp-mtblty}} assumes independence of features. This raises a valid concern about the effect of protecting immutable features in the presence of proxies that remain unprotected. We address this in @sec-discussion.
